A novel route to crystallization via viscoelastic phase separation

### 2015-10-19 science/crystal_gel_letter3.tex
Merged Mathieu's and sensei versions.
The working copy is crystal_gel_letter3.tex

In our experimental setup, the sample cell is put in contact with a reservoir of a salt solution through a semi-permeable membrane (see Methods 
on the details of our experiments).
At $t=0$, the increase in the concentration of salt ions progressively screens the Coulombic repulsion between colloidal particles,
which are then subject to attractive depletion forces, making the system thermodynamically unstable and leading to phase separation. 
%In the following we discuss all the stages of the formation of the crystal gels: phase separation, stress-driven agein and crystallization.
The experimental data are taken at two different volume fractions ($\phi\approx 0.13$ and $\phi\approx 0.30$) and for different values
of the polymer concentration, $c_p=0.38,0.48,0.57,1.07$ mg/g for $\phi\approx 0.30$ and $c_p=0.82,1.36$ mg/g for $\phi\approx 0.13$.
In Fig.~\ref{fig:phase_separation}A, we superimpose the state points with the phase diagram, which we calculated from generalised free volume theory~\cite{Fleer2008} using the experimental values for the size of the colloids, the radius of gyration of the polymer, and the polymer concentration (see Supplementary Information).
For each state point, scans were made at early times (every $10$~s, i.e. $4\tau_B$) and at the late stage (every $30$~s) of the gelation process.

@@

To gain additional insight, in Fig.~\ref{fig:phase_separation}D we plot the average local volume fraction, computed
through Voronoi diagrams, as a function of the number of neighbors. For all state points, we can see that the average volume fraction of a colloidal
particle increases with its number of neighbors. Approaching close packing (12 neighbors), we clearly see two families of curves. The state point
at $\phi\approx 0.30$ and $c_p=0.38$ mg/g reaches close packing at a volume fraction of around $\sim 68\%$, which is in reasonable agreement with the
volume fraction of the stable crystal phase~\cite{ilett1995phase}. All other state points, instead, reach close packing at a volume fraction $\sim 64\%$, which is
indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. 
%(\textbf{the absolute values of these volume fraction are not totally convincing, probably because of a not optimal estimate of the size of the colloids. We should run the code of Mathieau with size detection}). 
This result clearly indicates that the state point at $\phi\approx 0.30$ and $c_p=0.38$ mg/g follows a different arrest mechanism, in which phase separation
is arrested by crystallization and not by glassiness. Although our estimation of the absolute volume fraction may involves errors of a few \%, the trend should be robust.

@@

Next we show in Fig.~\ref{fig:transitions}B the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.30$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred. The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform in small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to appear in the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapor pressure of the liquid phase compared to the crystalline phase. After the onset
of a steady-state gas population, the crystallization proceeds by adding surface particles from the gas phase to the small crystallites. To confirm this process we
examine in Fig.~\ref{fig:transitions}C the transfer matrix probabilities $P_{\alpha\beta}$, where $\alpha$ and $\beta$ are each of the 
phases, gas, liquid and crystal (which includes hcp, fcc and surface particles). Liquid particles undergo two processes: on one
hand they transform to crystallites growing inside the liquid branches of the network, and also increasingly transform to gas particles.
Gas particles first undergo a transition to liquid particles as the final stage of the liquid-gas phase separation, but once
crystalline nuclei outgrow the liquid pockets in which they are formed, the probability for the transition from gas to liquid
abruptly decreases. Instead, gas particles start depositing on the crystal, and at later stages the pace of the gas-to-crystal transformation (sublimation)
is larger than the one of the liquid-to-crystal one.

>>>>

In our experimental setup, the sample cell is put in contact with a reservoir of a salt solution through a semi-permeable membrane (see Methods 
on the details of our experiments).
At $t=0$, the increase in the concentration of salt ions within a few Brownian times ($\tau_B$) screens the Coulombic repulsion between colloidal particles,
which are then subject to attractive depletion forces, making the system thermodynamically unstable and leading to phase separation. 
%In the following we discuss all the stages of the formation of the crystal gels: phase separation, stress-driven agein and crystallization.
The experimental data are taken at two different volume fractions ($\phi\approx 0.13$ and $\phi\approx 0.30$) and for different values
of the polymer concentration, $c_p=0.38,0.48,0.57,1.07$ mg/g for $\phi\approx 0.30$ and $c_p=0.82,1.36$ mg/g for $\phi\approx 0.13$.
In Fig.~\ref{fig:phase_separation}A, we superimpose the state points with the phase diagram, which we calculated from generalised free volume theory~\cite{Fleer2008} using the experimental values for the size of the colloids, the radius of gyration of the polymer, and the polymer concentration (see Supplementary Information).
For each state point, scans were made at early times (every $10$~s, i.e. $4\tau_B$) and at the late stage (every $30$~s) of the gelation process.

@@

To gain additional insight, in Fig.~\ref{fig:phase_separation}D we plot the average local volume fraction, computed
through Voronoi diagrams, as a function of the number of neighbors. For all state points, we can see that the average volume fraction of a colloidal
particle increases with its number of neighbors. Approaching close packing (12 neighbors), we clearly see two families of curves. The state point
at $\phi\approx 0.30$ and $c_p=0.38$ mg/g reaches close packing at a volume fraction of around $\sim 68\%$, which is in reasonable agreement with the
volume fraction of the stable crystal phase~\cite{ilett1995phase}. All other state points, instead, reach close packing at a volume fraction $\sim 64\%$, which is
indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. 
%(\textbf{the absolute values of these volume fraction are not totally convincing, probably because of a not optimal estimate of the size of the colloids. We should run the code of Mathieau with size detection}). 
This result clearly indicates that the state point at $\phi\approx 0.30$ and $c_p=0.38$ mg/g follows a different arrest mechanism, in which phase separation
is arrested by crystallization and not by glassiness. Although our estimation of the absolute volume fraction may involves systematic errors of a few \%, the trend should be robust~\cite{Poon2012}.

@@

Next we show in Fig.~\ref{fig:transitions}B the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.30$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred. The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform in small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapor pressure of the liquid phase compared to the crystalline phase. After the onset
of a steady-state gas population, the crystallization proceeds by adding surface particles from the gas phase to the small crystallites. To confirm this process we
examine in Fig.~\ref{fig:transitions}C the transfer matrix probabilities $P_{\alpha\beta}$, where $\alpha$ and $\beta$ are each of the 
phases, gas, liquid and crystal (which includes hcp, fcc and surface particles). Liquid particles undergo two processes: on one
hand they transform to crystallites growing inside the liquid branches of the network, and also increasingly transform to gas particles.
Gas particles first undergo a transition to liquid particles as the final stage of the liquid-gas phase separation, but once
crystalline nuclei outgrow the liquid pockets in which they are formed, the probability for the transition from gas to liquid
abruptly decreases. Instead, gas particles start depositing on the crystal, and at later stages the pace of the gas-to-crystal transformation (sublimation)
is larger than the one of the liquid-to-crystal one.
